---
title: "Lab 06 Replication"
author: "Branson Fox, BA and Christopher Prener, PhD"
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: github_document
---

## Introduction
This notebook replicates the results of lab 06

## Dependencies
This notebook requires the following packages to load and manipulate our data.

```{r load-packages}
# tidyverse packages
library(readr) # import tabular data
library(dplyr) # data manipulation

# spatial packages
library(sf)    # methods for spatial data
library(mapview) # interactive map preview
library(tigris) # TIGER/Line Data

# other packages
library(here)  # file path management

# functions
source(here("source", "map_breaks.R"))
```

## Load Data
These are the data we need to complete this lab.
```{r load-data}
# spatial 
county_bnd <-st_read(here("data","US_BOUNDARY_Counties","US_BOUNDARY_Counties.shp"))

# tabular
capitals <- read_csv(here("data","stateCapitals.csv"))
health_ins <- read_csv(here("data","USHealth","US_HEALTH_noIns.csv"))
```

## Part 1
### Question 1
We'll start by projecting the capitals data we loaded earlier:

```{r p1-q1-a}
capitals_sf <- st_as_sf(capitals, coords = c(x = "longitude", y = "latitude"), crs = 4269)
```

Then, we'll check that we projected properly using `mapview`:

```{r p1-q1-a2}
mapview(capitals_sf)
```

Next, we'll save the data as a shapefile using the `st_write()` function.

```{r save shapefile}
st_write(capitals_sf, dsn = here("data", "cleanData", "stateCapitals.shp"), delete_dsn = TRUE)
```

### Part B
Next, we'll work with the county boundary and health insurance data. First, we want to join these data. `GEOID` appears to be a common variable among them, but it is character in one data object and numeric in the other. To fix this, we can `mutate` one class to match the other.

```{r mutate-geoid}
health_ins <- mutate(health_ins, GEOID = as.numeric(GEOID))
```

Now we can table join as we have before. We'll subset down to the bare minimum of counties before we do that, however.

```{r table-join}
## subset columns
county_bnd <- select(county_bnd, GEOID)

## join
county_health <- left_join(county_bnd, health_ins, by = "GEOID")
```

### Part C
We want to subset our data to remove the observations with missing data. Since we know that `-1` denotes missing data, we'll `filter` for observations where `noIns` is `>=0`

```{r filter missing}
county_health <- filter(county_health, noIns >= 0)
```

### Part D
We're almost ready to export the data, but first we need to check the projection. We can do this with `st_crs()`

```{r check projection}
st_crs(county_health)
```

We find that there is not the correct coordinate system. We'll use `st_transform()` to re-project the data to NAD 1983:

```{r set crs}
county_health <- st_transform(county_health, crs = 4269)
```

Finally, we are ready to export:

```{r export shapefile2}
st_write(county_health, dsn = here("data", "cleanData", "countyHealth.shp"), delete_dsn = TRUE)
```

### Part E
Next, we want to assign map breaks to our data before we get to mapping. We'll also do this before we subset our data so that all three maps use the same breaks.

```{r map-breaks}
county_health <- map_breaks(county_health, var = "noIns", newvar = "map_breaks", classes = 5, style = "fisher")
```

### Part F
Next, we'll download state boundary data from Tiger/Line using the `tigris` package and prepare it for mapping.

```{r get-states}
## download and remove territories
states <- states() %>%
  filter(NAME %in% c("United States Virgin Islands", "Commonwealth of the Northern Mariana Islands", "Guam",
                     "American Samoa", "Puerto Rico") == FALSE)

## create lower 48, AK, and HI
lower48_boundary <- filter(states, NAME %in% c("Alaska", "Hawaii") == FALSE)
ak_boundary <- filter(states, NAME == "Alaska")
hi_boundary <- filter(states, NAME == "Hawaii")
```

Now we have data that are specific to the three maps we'll be making.

### Part G
We'll also remove Alaska and Hawaii from our other data sources as well:

```{r modify-data}
# capitals data
lower48_capitals <- filter(capitals, name %in% c("Alaska", "Hawaii") == FALSE)
ak_capital <- filter(capitals, name == "Alaska")
hi_capital <- filter(capitals, name == "Hawaii")

# health insurance data
lower48_health <- filter(cnty_health, state %in% c("Alaska", "Hawaii") == FALSE)
ak_health <- filter(cnty_health, state == "Alaska")
hi_health <- filter(cnty_health, state == "Hawaii")
```

Now we have our other data prepared for mapping as well.
