---
title: "Lab 06 Replication"
author: "Branson Fox, BA and Christopher Prener, PhD"
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: github_document
---

## Introduction
This notebook replicates the results of lab 06

## Dependencies
This notebook requires the following packages to load and manipulate our data.

```{r load-packages}
# tidyverse packages
library(readr) # import tabular data
library(dplyr) # data manipulation

# spatial packages
library(sf)    # methods for spatial data
library(mapview) # interactive map preview
library(tigris) # TIGER/Line Data

# other packages
library(here)  # file path management
```

## Load Data
These are the data we need to complete this lab.
```{r load-data}
# spatial 
county_bnd <-st_read(here("data","US_BOUNDARY_Counties","US_BOUNDARY_Counties.shp"))

# tabular
capitals <- read_csv(here("data","stateCapitals.csv"))
health_ins <- read_csv(here("data","USHealth","US_HEALTH_noIns.csv"))
```

## Part 1
### Question 1
We'll start by projecting the capitals data we loaded earlier.
```{r p1-q1-a}
capitals_sf <- st_as_sf(capitals, coords = c(x = "longitude", y = "latitude"), crs = 4269)
```

Then, we'll check that we projected properly using `mpaview`
```{r p1-q1-a2}
mapview(capitals_sf)
```

Next, we'll save the data as a shapefile using the `st_write()` function.

```{r save shapefile}
st_write(capitals_sf, dsn = here("data", "cleanData", "stateCapitals.shp"), delete_dsn = TRUE)
```

### Part B
Next we'll work with the county boundary and health insurance data. First, we want to join these data. `GEOID` appears to be a common variable among them, but it is character in one data.frame and numeric in the other. To fix this, we can `mutate` one class to match the other.

```{r mutate class}
health_ins <- mutate(health_ins, GEOID = as.numeric(GEOID))
```

Now we can table join as usual.

```{r table join}
cnty_health <- left_join(county_bnd, health_ins, by = "GEOID")
```

### Part C
We want to subset our data to remove the observations with missing data. Since we know that $-1$ denotes missing data, we'll `filter` for observations where `noIns` is `>=0`

```{r filter missing}
cnty_health <- filter(cnty_health, noIns >= 0)
```

### Part D
We're almost ready to export the data, but first we need to check the projection. We can do this with `st_crs()`

```{r check projection}
st_crs(cnty_health)
```

We find that there is not the correct coordinate system. We'll use `st_transform()` to re-project the data to NAD 1983:

```{r set crs}
cnty_health <- st_transform(cnty_health, crs = 4269)
```

Finally, we are ready to export:

```{r export shapefile2}
st_write(cnty_health, dsn = here("data", "cleanData", "countyHealth.shp"), delete_dsn = TRUE)
```

## Part 2
### Question 2
Next, we'll download state boundary data from Tiger/Line using the `tigris` package and prepare it for mapping.

```{r get-states}
## download and remove territories
states <- states() %>%
  filter(NAME %in% c("United States Virgin Islands", "Commonwealth of the Northern Mariana Islands", "Guam",
                     "American Samoa", "Puerto Rico") == FALSE)

## create lower 48
lower48 <- filter(states, NAME %in% c("Alaska", "Hawaii") == FALSE)
```

We'll also remove Alaska and Hawaii from our other data