---
title: "Lab 3-1 Replication"
author: "Branson Fox, BA and Christopher Prener, PhD"
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: github_document
always_allow_html: true
---

## Introduction
This notebook replicates the results of lab 06

## Dependencies
This notebook requires the following packages to load and manipulate our data.

```{r load-packages}
# tidyverse packages
library(dplyr)          # data manipulation
library(ggplot2)        # static mapping
library(readr)          # import tabular data

# spatial packages
library(sf)             # methods for spatial data
library(mapview)        # interactive map preview
library(tigris)         # TIGER/Line Data

# other packages
library(here)           # file path management
library(RColorBrewer)   # color palettes

# functions
source(here("source", "map_breaks.R"))
```

## Load Data
These are the data we need to complete this lab.

```{r load-data}
# spatial 
county_bnd <-st_read(here("data","US_BOUNDARY_Counties","US_BOUNDARY_Counties.shp"))

# tabular
capitals <- read_csv(here("data","stateCapitals.csv"))
health_ins <- read_csv(here("data","USHealth","US_HEALTH_noIns.csv"))
```

## Part 1
### Section A
We'll start by projecting the capitals data we loaded earlier:

```{r p1-q1-a}
capitals_sf <- st_as_sf(capitals, coords = c(x = "longitude", y = "latitude"), crs = 4269)
```

Then, we'll check that we projected properly using `mapview`:

```{r p1-q1-a2}
mapview(capitals_sf)
```

Next, we'll save the data as a shapefile using the `st_write()` function.

```{r save shapefile}
st_write(capitals_sf, dsn = here("data", "cleanData", "stateCapitals.shp"), delete_dsn = TRUE)
```

### Section B
Next, we'll work with the county boundary and health insurance data. First, we want to join these data. `GEOID` appears to be a common variable among them, but it is character in one data object and numeric in the other. To fix this, we can `mutate` one class to match the other.

```{r mutate-geoid}
health_ins <- mutate(health_ins, GEOID = as.numeric(GEOID))
```

Now we can table join as we have before. We'll subset down to the bare minimum of counties before we do that, however.

```{r table-join}
## subset columns
county_bnd <- select(county_bnd, GEOID)

## join
county_health <- left_join(county_bnd, health_ins, by = "GEOID")
```

### Section C
We want to subset our data to remove the observations with missing data. Since we know that `-1` denotes missing data, we'll `filter` for observations where `noIns` is `>=0`

```{r filter missing}
county_health <- filter(county_health, noIns >= 0)
```

### Section D
We're almost ready to export the data, but first we need to check the projection. We can do this with `st_crs()`

```{r check projection}
st_crs(county_health)
```

We find that there is not the correct coordinate system. We'll use `st_transform()` to re-project the data to NAD 1983:

```{r set crs}
county_health <- st_transform(county_health, crs = 4269)
```

Finally, we are ready to export:

```{r export shapefile2}
st_write(county_health, dsn = here("data", "cleanData", "countyHealth.shp"), delete_dsn = TRUE)
```

### Section E
Next, we want to assign map breaks to our data before we get to mapping. We'll also do this before we subset our data so that all three maps use the same breaks.

```{r map-breaks}
county_health <- map_breaks(county_health, var = "noIns", newvar = "map_breaks", classes = 5, style = "fisher")
```

### Section F
Next, we'll download state boundary data from Tiger/Line using the `tigris` package and prepare it for mapping.

```{r get-states}
## download and remove territories
states <- states() %>%
  filter(NAME %in% c("United States Virgin Islands", "Commonwealth of the Northern Mariana Islands", "Guam",
                     "American Samoa", "Puerto Rico") == FALSE)

## create lower 48, AK, and HI
lower48_boundary <- filter(states, NAME %in% c("Alaska", "Hawaii") == FALSE)
ak_boundary <- filter(states, NAME == "Alaska")
hi_boundary <- filter(states, NAME == "Hawaii")
```

Now we have data that are specific to the three maps we'll be making.

### Section G
We'll also remove Alaska and Hawaii from our other data sources as well:

```{r modify-data}
# capitals data
lower48_capitals <- filter(capitals_sf, name %in% c("Alaska", "Hawaii") == FALSE)
ak_capital <- filter(capitals_sf, name == "Alaska")
hi_capital <- filter(capitals_sf, name == "Hawaii")

# health insurance data
lower48_health <- filter(county_health, state %in% c("Alaska", "Hawaii") == FALSE)
ak_health <- filter(county_health, state == "Alaska")
hi_health <- filter(county_health, state == "Hawaii")
```

Now we have our other data prepared for mapping as well.

## Part 2
### Question 2
We'll use the [Albers Contiguous USA](https://spatialreference.org/ref/esri/usa-contiguous-albers-equal-area-conic/) projected coordinate system, whose EPSG code is not a part of `R`'s ecosystem.

```{r p2-q2}
## state capitals
lower48_capitals <- st_transform(lower48_capitals, crs = "ESRI:102003")

## state boundary
lower48_boundary <- st_transform(lower48_boundary, crs = "ESRI:102003")

## health insurance data
lower48_health <- st_transform(lower48_health, crs = "ESRI:102003")
```

Other options include the more general [Albers North America](https://spatialreference.org/ref/esri/102008/) coordinate system and the [Lambert North America](https://spatialreference.org/ref/esri/north-america-lambert-conformal-conic/) coordinate system.

### Question 3
Next, we'll use these data to make a map of the lower 48 states.

```{r p2-q3}
p1 <- ggplot() +
  geom_sf(data = lower48_boundary, fill = "#ffffff", color = NA) +
  geom_sf(data = lower48_health, mapping = aes(fill = map_breaks), lwd = .4) +
  geom_sf(data = lower48_boundary, fill = NA, color = "#2a2a2a", size = .7) +
  geom_sf(data = lower48_capitals, shape = 18) +
  scale_fill_brewer(palette = "RdPu", name = "% of Residents\nwithout Insurance") +
  labs(
    title = "Insurance Rates by County",
    subtitle = "Contiguous United States",
    caption = "Data via the CDC and Christopher Prener, PhD"
  )

p1
```

### Question 4
With our map created, we'll save it to the `results/` folder:

```{r p2-q4}
ggsave(plot = p1, filename = here("results", "lower48_map.png"))
```

## Part 3
### Question 5
We'll use the [Albers Alaska](https://spatialreference.org/ref/epsg/3467/) projected coordinate system, which is specifically designed for state-level mapping in Alaska.

```{r p3-q5}
## state capitals
ak_capital <- st_transform(ak_capital, crs = 3467)

## state boundary
ak_boundary <- st_transform(ak_boundary, crs = 3467)

## health insurance data
ak_health <- st_transform(ak_health, crs = 3467)
```

The other option is the more general [Albers North America](https://spatialreference.org/ref/esri/102008/) coordinate system, though it is preferable to use state-specific coordinate systems for places like Alaska.

### Question 6
Next, we'll use these data to make a map of the Alaska. One tweak to our workflow that go above and beyond what we've covered in class is included here: I've used `scale_fill_manual()` to specify the three darkest colors form our map above. If we use the default skill from class, `scale_fill_brewer()`, it will select the first three color values from this palette.

```{r p3-q6}
p2 <- ggplot() +
  geom_sf(data = ak_boundary, fill = "#ffffff", color = NA) +
  geom_sf(data = ak_health, mapping = aes(fill = map_breaks), lwd = .4) +
  geom_sf(data = ak_boundary, fill = NA, color = "#2a2a2a", size = .7) +
  geom_sf(data = ak_capital, shape = 18, size = 4) +
  scale_fill_manual(values = brewer.pal(name = "RdPu", n = 5)[3:5],
                    name = "% of Residents\nwithout Insurance") +
  labs(
    title = "Insurance Rates by County",
    subtitle = "Alaska Focus",
    caption = "Data via the CDC and Christopher Prener, PhD"
  )

p2
```

This gets us reasonably close to a production quality map, though some additional tweaks, such as showing specific counties that have missing data, might make it a nicer final product.

### Question 7
With our map created, we'll save it to the `results/` folder:

```{r p3-q7}
ggsave(plot = p2, filename = here("results", "alaska_map.png"))
```

## Part 4
### Question 8
We'll use the [Albers Hawaii](https://spatialreference.org/ref/esri/102007/) projected coordinate system, which is specifically designed for state-level mapping in Hawaii.

```{r p4-q8}
## state capitals
hi_capital <- st_transform(hi_capital, crs = "ESRI:102007")

## state boundary
hi_boundary <- st_transform(hi_boundary, crs = "ESRI:102007")

## health insurance data
hi_health <- st_transform(hi_health, crs = "ESRI:102007")
```

The other option is a variation of the UTM system that is specific to Hawaii - [Hawaii UTM Zone 4N](https://spatialreference.org/ref/sr-org/9086/).

### Question 9
Next, we'll use these data to make a map of the Hawaii. 

```{r p4-q9}
p3 <- ggplot() +
  geom_sf(data = hi_health, mapping = aes(fill = map_breaks), color = "#2a2a2a", size = .7) +
  geom_sf(data = hi_capital, shape = 18, size = 4) +
  scale_fill_brewer(palette = "RdPu", name = "% of Residents\nwithout Insurance") +
  labs(
    title = "Insurance Rates by County",
    subtitle = "Hawaii Focus",
    caption = "Data via the CDC and Christopher Prener, PhD"
  )

p3
```

After playing around with this map, I decided not to use the `hi_boundary` data because it added a number of additional islands that we do not have data for.

### Question 10
With our map created, we'll save it to the `results/` folder:

```{r p4-q10}
ggsave(plot = p3, filename = here("results", "hawaii_map.png"))
```
